Bootstrapping Multi-atlas Hippocampal Subfield Segmentation by using Multiple Automatically Generated Templates

### 2013-09-02 paper.Rnw
The rest of Julie's edits

% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% TODO: 
%  - check tense: we are is have has 
%  - update montages with TAL coordinates
%  - add MAGeT brain diagram, and remove pseudocode
% Sweave/Knitr init
<<setup, include=FALSE, cache=FALSE>>=
library(knitr)
opts_chunk$set(echo = FALSE, 
               message = FALSE, 
               warning = FALSE, 
               fig.align = 'center', 
               tidy = FALSE, 
               comment = NA, 
               cache = TRUE,
               fig.width = 6, 
               fig.height = 6)
library(ggplot2)
library(Hmisc)
library(reshape2)
library(ADNIMERGE)
library(plyr)
options(digits=3)
theme_set(theme_bw(base_size = 12))

@@

Figure \ref{alg:MAGeT} describes the \mb algorithm in pseudocode. Source code 
for \mb can be found at \url{http://github.com/pipitone/MAGeTbrain}.

\begin{algorithm}
  \scriptsize
  \caption{Pseudocode for the \mb algorithm}
  \label{alg:MAGeT}
  \begin{algorithmic}
    \Function{BasicMultiAtlasSegmentation}{Templates, Subjects}
      \ForAll{$target$}
        \ForAll{$template$}
          \State propagate all labels for template to target space
          \State store target labels
        \EndFor
        \State fuse target labels
      \EndFor
    \EndFunction
    \\
    \Function {MAGeTBrain}{Subjects, Atlases, n}
      \For{$i = 1 \to n$}
        \State choose a target to be used as a template
        \State propagate labels from each atlas to template space
        \State store the template with all of its labels
      \EndFor
      \State MultiAtlas(Templates, Subjects)
    \EndFunction
  \end{algorithmic}
\end{algorithm}

@@

Figure \ref{fig:ADNI-subfields} shows a qualitative comparison of \mb and 
SNT \gs hippocampal segmentations for 10 randomly selected subjects in 
each disease category, and illustrates some of the common errors found during 
visual inspection. Mostly frequently, we find \mb improperly includes the vestigial
hippocampal sulcus and, although not anatomically incorrect, \mb under-estimates 
the hippocampal body in comparison to the SNT \gs segmentation.

@@

\begin{figure}
  \begin{centering}
    \includegraphics[width=6in]{figure/ADNI1_SNT_MB_montage/montage.pdf}
  \end{centering}
  \caption{SNT \gs and \mb segmentations for 30 ADNI1 subjects (10
  subjects randomly selected from each disease category in the subject pool used
  in Experiment 1). Sagittal slices are shown for each unlabelled T1-weighted
  anatomical image. SNT \gs labels appear in green, and \mb labels appear in
  blue. Noted are examples of common segmentation idiosyncrasies: 
  {\em (a)} over-estimation of hippocampal head and  
  {\em (b)} translated SNT \gs segmentation; 
  {\em (c)} under-estimation of hippocampal body and
  {\em (d)} improper inclusion of the vestigial hippocampal sulcus by \mb.}
  \label{fig:ADNI-subfields}
\end{figure}

@@

In this manuscript we have presented the implementation and validation of the
\mb framework -- a methodology that requires very few input atlases in
order to provide accurate and reliable segmentations.  We have demonstrated that
our methodology robustly provides accurate and consistent segmentations in
populations with different ageing and neuropsychiatric characteristics.
Further, we have demonstrated that algorithmic performance is not dependent on a
single definition of the hippocampus but is effective with differing hippocampal
definitions \citep{Winterburn2013,Pruessner2000,Hsu2002}.  Finally, we also demonstrate
that \mb provides accurate automatic identification of the hippocampal
subfields despite contrast and resolution limitations in standard T1-weighted
image volumes.

@@

Another concern is the moderate-to-small improvement observed in \mb in
comparison to multi-atlas segmentation when using the same number of atlases.
The actual benefit in using \mb is consistency of the labelling regardless of
atlas or template choice. This is an important consideration that few have
touched on previously. The 10-fold Monte Carlo leave-one-out cross-validation
that we present in Experiment 1 is amongst one of the most stringent performed
in the multi-atlas/segmentation literature. To the best of our knowledge, other
groups using ADNI data for validation do at most a single round of
leave-one-out-validation (Table \ref{tab:other-methods}), aside from
\citep{Wang2011} who perform a similar validation as us). The thoroughness of
our validation suggests that our results are reflective of a true average over
the choice of parameter settings and are independent of atlas or template
choice (provided the input atlases are properly segmented.

@@

In comparison to other methodologies in the field \mb performs favourably. Table
\ref{tab:other-methods} surveys some of the most recent reported kappa values on
the ADNI dataset using the same SNT labels for populating their template library
and as gold standards for evaluation.  While it is difficult to compare
segmentation results across studies, gold-standards, evaluation metrics, and
algorithms it is worth noting that the methods summarized in Table
\ref{tab:other-methods} require more atlases (between 16-55) than our \mb
implementation with the Winterburn atlases \citep{Winterburn2013}.

There are some important differences between our method and these specific
methods.  Others have reported the difficulty with mis-registrations in
candidate segmentation (i.e. segmentations generated that are then input in the
voxel-voting procedure \citep{Collins2010}). The work of \citet{Leung2010}
tackles this problem by using an intensity threshold that is estimated
heuristically at the time of segmentation (this work also reports some of the
highest kappas for the segmentation of ADNI data).  While this method is
effective for the ADNI dataset (which is partially homogenized with respect to
image acquisition and pre-processing), it is unclear if this type of heuristic
is applicable to other datasets.  In all cases, these methods require more
atlases than our implementation with the Winterburn atlases.
\citet{Lotjonen2010} achieve highly accurate segmentation but correct their
segmentations using classifications derived using an expectation maximization
framework. In their initial work, \citet{Chupin2009} develop their probabilistic
methodology using a cohort of 8 healthy controls and 15 epilepsy patients, and
then use this method to segment an ADNI sample, with a hierarchical
experimentation protocol that involved \todo{complete this sentence}.  These
methods suggest that some post-processing of the final segmentations would
improve accuracy of the segmentation. While that may be true, there is little
consensus regarding how to achieve this.

@@

In comparison to the methods that we compared and were available through the
ADNI database (\freesurfer, \maper) and \fsl-FIRST, we find extremely variable
performance of all methods.  With the exception of \fsl all methods correlate
well with the SNT volumes provided in the ADNI database.  However, \freesurfer
and \fsl provide radically different definitions of the size of the hippocampus
in comparison to the other methods.  Further, when estimating bias of these
methods relative to SNT hippocampal volumes we see that large hippocampi are
over estimated while small hippocampi are under estimated. By comparison, \mb
and MAPER are far more conservative in volume estimation, suggesting these
methods may be better suited for estimating true-positives. However, in this
analysis we have only compared methods by total hippocampal volume, and so more
work is needed to understand the full extent to which these methods differ.

Finally, we have also demonstrated that our algorithmic framework is
appropriate for the segmentation of hippocampal subfields in standard
T1-weighted data.  Subfield segmentation is a burgeoning topic in the
literature although very few automated methods are available for the
segmentation of 3T data \citep{Yushkevich2009,VanLeemput2009}. However, the
FreeSurfer implementation of subfield segmentation \todo{complete}. While
recent work demonstrates that subfield segmentations can be used for
classification of AD, MCI, and NC, there has been no explicit validation of the
methodology based on accuracy or precision.  Although the initial hippocampal
subfield segmentations from the Yushkevich group have been demonstrated to work
on the ADNI population, there has also been no validation of their work.  In
addition, their work requires some manual initialization to properly function.
Our work demonstrates that we can reliably identify the CA1, subiculum, and CA4
dentate with only modest amounts of error.  The fact that CA2/CA3 and molecular
layers cannot be reasonably identified should not be surprising as these are
extremely thin and spatially convoluted regions that originally required
high-resolution MRI for identification.  It is likely that extents of these
regions are well below the resolution offered by standard T1-weighted images
and, in fact, many manual segmentation methodologies do not attempt to parse
these regions either \citep{Wisse2012, Mueller2009}.

>>>>

% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% TODO: 
%  - check tense: we are is have has 
%  - add MAGeT brain diagram, and remove pseudocode
% Sweave/Knitr init
<<setup, include=FALSE, cache=FALSE>>=
library(knitr)
opts_chunk$set(echo = FALSE, 
               message = FALSE, 
               warning = FALSE, 
               fig.align = 'center', 
               tidy = FALSE, 
               comment = NA, 
               cache = TRUE,
               fig.width = 6, 
               fig.height = 6)
library(ggplot2)
library(Hmisc)
library(reshape2)
library(ADNIMERGE)
library(plyr)
options(digits=3)
theme_set(theme_bw(base_size = 12))

@@

Figure \ref{fig:MAGeT} illustrates the \mb algorithm graphically. Source code
for \mb can be found at \url{http://github.com/pipitone/MAGeTbrain}.

\begin{figure}
  \caption{The \mb algorithm \label{fig:MAGeT}}
\end{figure}

@@

Figure \ref{fig:ADNI-segmentations} shows a qualitative comparison of \mb and 
SNT \gs hippocampal segmentations for 10 randomly selected subjects in 
each disease category, and illustrates some of the common errors found during 
visual inspection. Mostly frequently, we find \mb improperly includes the vestigial
hippocampal sulcus and, although not anatomically incorrect, \mb under-estimates 
the hippocampal body in comparison to the SNT \gs segmentation.

@@

\begin{figure}
  \begin{centering}
    \includegraphics[width=6in]{figure/ADNI1_SNT_MB_montage/montage.pdf}
  \end{centering}
  \caption{SNT \gs and \mb segmentations for 30 ADNI1 subjects (10
  subjects randomly selected from each disease category in the subject pool used
  in Experiment 1). Sagittal slices are shown for each unlabelled T1-weighted
  anatomical image. SNT \gs labels appear in green, and \mb labels appear in
  blue. Noted are examples of common segmentation idiosyncrasies: 
  {\em (a)} over-estimation of hippocampal head and  
  {\em (b)} translated SNT \gs segmentation; 
  {\em (c)} under-estimation of hippocampal body and
  {\em (d)} improper inclusion of the vestigial hippocampal sulcus by \mb.}
  \label{fig:ADNI-segmentations}
\end{figure}

@@

In this manuscript we have presented the implementation and validation of the
\mb framework -- a methodology that requires very few input atlases in
order to provide accurate and reliable segmentations.  Experiment 1 establishes
\mb as effective, and as comparable to multi-atlas segmentation. This experiment
also rigorously characterizes the behaviour of \mb whole hippocampal
segmentation under various parameter settings allowing us to choose an optimal
setting for subsequent experiments. Experiment 2 demonstrates the accuracy of
\mb hippocampal subfield identification despite contrast and resolution
limitations in standard T1-weighted image volumes.  Experiments 2 and 3 validate
\mb whole hippocampal segmentation accuracy and consistency on populations with
different ageing and neuropsychiatric characteristics.  Furthermore, taken
together, these experiments demonstrate that algorithmic performance is not
dependent on a single definition of the hippocampus but is effective with
differing hippocampal definitions \citep{Winterburn2013,Pruessner2000,Hsu2002}.

@@

Another concern is the moderate-to-small improvement observed in \mb in
comparison to multi-atlas segmentation when using the same number of atlases.
The actual benefit in using \mb is consistency of the labelling regardless of
atlas or template choice. This is an important consideration that few have
touched on previously. The 10-fold Monte Carlo leave-one-out cross-validation
that we present in Experiment 1 is amongst one of the most stringent performed
in the multi-atlas/segmentation literature. To the best of our knowledge, other
groups using ADNI data for validation do at most a single round of
leave-one-out-validation (Table \ref{tab:other-methods}), aside from
\citep{Wang2011} who perform a similarily rigorous validation as us). The
thoroughness of our validation suggests that our results are reflective of a
true average over the choice of parameter settings and are independent of atlas
or template choice (provided the input atlases are consistently segmented).

On that note, it is worth observing that we found the SNT segmentations to be
poor in general. As previously noted in Figure \ref{fig:ADNI-segmentations}, an
expert manual rater (JW) identified many inconsistencies: occurances of over-
and under-estimation, as well as misalignments of the entire segmentation
volume.  Therefore, the use of the SNT segmentations as gold standards in
Experiments 1 and 4 presents the possibility that the DSC scores and
correlations we obtained are lower than what would be found with a more accurate
and consistent gold standard.

@@

In comparison to other methodologies in the field \mb performs favourably. Table
\ref{tab:other-methods} surveys some of the most recent reported DSC values
reported on ADNI dataset, using SNT segmentations for the atlas library and as
gold standards for evaluation.  While it is difficult to compare segmentation
results across studies, gold standards, evaluation metrics, and algorithms it is
worth noting that the methods summarized require more atlases (between 16-55)
than our \mb implementation with the Winterburn atlases \citep{Winterburn2013}.

There are some important differences between our method and these specific
methods. Others have reported the difficulty with mis-registrations in
candidate segmentation (i.e. segmentations generated that are then input in the
voxel-voting procedure \citep{Collins2010}). The work of \citet{Leung2010}
tackles this problem by using an intensity threshold that is estimated
heuristically at the time of segmentation (this work also reports some of the
highest DSC scores for the segmentation of ADNI data). While this method is
effective for the ADNI dataset (which is partially homogenized with respect to
image acquisition and pre-processing), it is unclear if this type of heuristic
is applicable to other datasets. In all cases, these methods require more
atlases than our implementation with the Winterburn atlases.
\citet{Lotjonen2010} achieve highly accurate segmentation but correct their
segmentations using classifications derived using an expectation maximization
framework. In their initial work, \citet{Chupin2009} develop their probabilistic
methodology using a cohort of 8 healthy controls and 15 epilepsy patients, and
then use this method to segment an ADNI sample, with a hierarchical
experimentation protocol. These methods suggest that some post-processing of the
final segmentations would improve accuracy of the segmentation. While that may
be true, there is little consensus regarding how to achieve this.

@@

Among the automated segmentation methods we compared in this paper
(\freesurfer, \maper, \fsl-FIRST), we find extremely variable
performance of all methods.  With the exception of \fsl all methods correlate
well with the SNT volumes provided in the ADNI database.  However, \freesurfer
and \fsl provide radically different definitions of the size of the hippocampus
in comparison to the other methods.  Further, when estimating bias of these
methods relative to SNT \gs hippocampal volumes we see that large hippocampi are
over estimated while small hippocampi are under estimated. By comparison, \mb
and MAPER are far more conservative in volume estimation, suggesting these
methods may be better suited for estimating true-positives, especially in
neurodegenerative disease subjects featuring smaller overall hippocampi.
However, in this analysis we have only compared methods by total hippocampal
volume, and so more work is needed to understand the full extent to which these
methods differ.

Finally, we have also demonstrated that our algorithmic framework is
appropriate for the segmentation of hippocampal subfields in standard
T1-weighted data.  Subfield segmentation is a burgeoning topic in the
literature although very few automated methods are available for the
segmentation of 3T data \citep{Yushkevich2009,VanLeemput2009}. However, the
FreeSurfer implementation of subfield segmentation \todo{complete}. While
recent work demonstrates that subfield segmentations can be used for
classification of AD, MCI, and CN, there has been no explicit validation of the
methodology based on accuracy or precision.  Although the initial hippocampal
subfield segmentations from the Yushkevich group have been demonstrated to work
on the ADNI population, there has also been no validation of their work.  In
addition, their work requires some manual initialization to properly function.
Our work demonstrates that we can reliably identify the CA1, subiculum, and
CA4/dentate with only modest amounts of error.  The fact that CA2/CA3 and
molecular layers cannot be reasonably identified should not be surprising as
these are extremely thin and spatially convoluted regions that originally
required high-resolution MRI for identification.  It is likely that the extents of
these regions are well below the resolution offered by standard T1-weighted
images and, in fact, many manual segmentation methodologies do not attempt to
parse these regions either \citep{Wisse2012, Mueller2009}.
